Statistically-estimated tree composition for the northeastern United States \\ at the time of European settlement

### 2015-04-19 paper.tex
added plos_one latex files and some minor edits to the paper

\title{Statistically-estimated tree composition for the northeastern United
States \\
at the time of European settlement}
\maketitle
\begin{quote}
Christopher J. Paciorek, Andrew Thurman, Simon J. Goring, Charlie
V. Cogbill, \\
John W. Williams, David J. Mladenoff, Jody A. Peters, Jun Zhu, \\
and Jason S. McLachlan {[}author order to be discussed{]}\\
{[}please tell me if you do/don't want your middle initial{]}
\end{quote}
TODO:

full edit

properly cite Goring stuff

@@

\begin{abstract}
We present a data product of the estimated composition of tree taxa
at the time of European settlement of the United States and the statistical
methodology used to produce the product. Composition is defined as
the proportion of stems larger than approximately X cm diameter at
breast height {[}Charlie? Simon? is X roughly 10?{]} in 22 taxonomic
groupings, generally at the genus level. The data come from settlement
survey records that provide raw data that are transcribed and then
aggregated spatially, giving count data. The domain is divided into
two regions, western (Indiana through Minnesota) and eastern (Ohio
to Maine). Public Land Survey point data in the western region are
aggregated to a regular 8 km grid, while data in the eastern region,
from Town Proprietor Surveys is aggregated at the township level.
The product is based on a Bayesian statistical model fit to the count
data that estimates composition on a regular 8 km grid. The statistical
model allows us to estimate composition at locations with no data
and to smooth over noise caused by limited counts in locations with
data. Critically, it also allows us to quantify uncertainty in our
composition estimates. We expect this data product to be useful for
understanding the state of vegetation in the northeastern United States
prior to large-scale European settlement. In addition to specific
regional questions, the data product can also serve as a baseline
against which to investigate how forests and ecosystems change after
intensive settlement.
\end{abstract}

@@

Historical datasets provide critical context to understand forest
ecology. Historical datasets allow researchers to define 'baseline'
conditions in conservation management, to understand ecosystem processes
at decadal and centennial scales, and, particularly in regions with
widespread land use change, to understand the extent to which forest
conversion and regeneration differ from the original forest cover.

Euro-American settlement and subsequent land use change occured in
a time transient fashion across North America, and surveys of vegetation
followed a similar pattern. Early surveys (from 1620 until 1825) in
the northeastern United States were areal surveys at a township level
\citep{cogbill2002forests,thompson2013four}, which we call the Town
Proprietor Survey (TPS). These surveys reporting percent composition
of major tree species using sometimes inconsistent common names. Later
surveys, after the establishment of the U.S. Public Land Survey System
(PLS) - from 1832 to 1907 - were point estimates along a regular grid,
with 1 mile spacing \citep{schulte2001original,bourdo1956review,goring2015composition}.
Survey instructions during the PLS varied through time, and by point
type. This required the application of spatially varying correction
factors \citep{cogbill2015corrections,goring2015composition} to accurately
assess stem density, basal area and biomass from the early settlement
records.

Logging, agriculture and abandonment has left an indelible mark on
forests in the northeastern United States \citep{foster1998land,goring2015composition,rhemtulla2009legacies,thompson2013four},
however most studies have studied these effects at the state level
or below \citep{rhemtulla2009historical,friedman2005regional}, and
with varying spatial resolution, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
provide a dataset across the upper Midwest that is resolved to an
8 x 8 km grid size, providing broad spatial coverage at a meso-spatial
scale that can be compared to the Forest Inventory and Analysis products
\citet{gray2012forest}. The limitation of this dataset is that it
provides only estimates of variance within cell for PLS data in the
western region, and no township estimates of error for regions surveyed
as part of the TPS.

Properly assessing uncertainty in ecological data is imperative to
understanding and modelling ecological processes \citep{cressie2009accounting}.
In this way, a model that can account for the spatial structure of
the underlying PLSS and TPS data, and provide reliable estimates of
uncertainty across the northeastern United States provides an enormously
valuable tool for researchers interested in the ecological structure
and function of forests at longer time scales.

@@

The raw data are obtained from survey records collated from across
the northeastern U.S. by a number of researchers. For the states of
Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
subdomain), data are available at survey point locations and have
been aggregated to a regular 8 km grid in the Albers projection. (Note
that for Indiana and Illinois, at the moment trees are associated
with township centroids and then assigned to 8 km grid cells based
on the centroid but in the near future we will have point locations
available for each tree.) For the states of Ohio, Pennsylvania, New
Jersey, New York and the six New England states (the eastern subdomain),
data are aggregated at the township level. There is also data from
a single township in Quebec and a single township in northern Delaware.
Data are essentially complete in Minnesota, Wisconsin and Michigan
but data in Illinois and Indiana represent a sample of the full set
of grid cells, with survey record transcription ongoing. Data for
the remaining states are a subset of the full set of townships. Fig.
\ref{fig:domain} {[}Fig. 1 - Lyx is having a numbering problem{]}
shows the domain, indicating the grid cells and townships with data.

@@

\caption{Spatial domain, with locations with data shown in gray. Locations
are grid cells in western portion and townships in eastern portion.
Grid cells completely covered in water are white (e.g., a few locations
in Minnesota and Wisconsin).}
\end{figure}

Note that surveys occurred over a period of more than 200 years as
European colonists (before U.S. independence) and the United States
settled what is now the northeastern United States. Our estimates
are for the period of settlement represented by the survey data and
therefore are time-transgressive; they do not represent any single
point in time across the domain, but rather the state of the landscape
at the time just before settlement occurred {[}get ref from Charlie
or Jason's historian pal{]}

Extensive details on the data are available in \cite{Gori:etal:2015}
and the raw data are available at XXX. The aggregation into taxonomic
groups is primarily at the genus level, but in some cases of monospecific
genera, is at the species level. We model the following 22 taxa plus
an ``other hardwood'' category: Atlantic white cedar (Chamaecyparis
thyoides), Ash (Fraxinus spp.), Basswood (Tilia americana), Beech
(Fagus grandifolia), Birch (Betula spp.), Black gum/sweet gum (Nyssa
sylvatica and Liquidambar styraciflua), Cedar/juniper (what is in
here other than Juniperus and is the Juniperus only Juniperus virginiana?),
Cherry (Prunus spp.), Chestnut (Castanea dentata), Dogwood (Cornus
spp.), Elm (Ulmus spp. or Ulmus americana???) , Fir (Abies spp.),
Hemlock (Tsuga canadensis), Hickory (Carya spp.), Ironwood (Carpinus
caroliniana and Ostrya virginiana (anything else?)), Maple (Acer spp.),
Oak (Quercus spp.), Pine (Pinus spp.), Poplar/tulip poplar (Populus
spp. and Liriodendron tulipifera), Spruce (Picea spp.), Tamarack (Larix
laricina), Walnut (Juglans nigra). {[}Simon/Charlie to check my Latin
and whether I've left out anything that falls into a given common
name{]} Note that in several cases (e.g., black gum/sweet gum, ironwood,
poplar/tulip poplar, cedar/juniper), because of ambiguity in the common
tree names used by surveyors, a group represents trees from different
families and even orders. For the western subdomain we do not fit
statistical models for Atlantic white cedar and chestnut as these
have 0 and 6 trees present, respectively. The taxa grouped into other
hardwood are those for which fewer than roughly 2000 trees were present
in the dataset; however we include Atlantic white cedar explicitly
despite it only having 336 trees in the dataset.

{[}include description of size cutoff{]} {[}Simon, Jody mentioned
that ther are 402 trees in IN below \textasciitilde{}10 cm and at
least 43 in IL. Do you screen those out or do we just live with it?{]}

@@

Our domain is a rectangle covering all of the states, in the Albers
projection (NAD 1983 Albers Great Lakes and St. Lawrence), with the
rectangle split into 8 km cells, arranged in a 296 by 180 grid of
cells, with the centroid of the cell in the southeast corner located
at (-71000, 58000). For the modeling of the western subdomain we use
the western-most 146 by 180 grid of cells. For the modeling of the
eastern subdomain we use the eastern-most 180 by 180 grid of cells
and then omit 23 cells in the north and 17 cells in the south.

@@

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
based on the full population of all the trees in an areal region.
\end{enumerate}
The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition of the taxa at all of the grid cells. These samples
are the data product and can then be used in analyses. We also provide
the mean and standard deviation of the samples, which represent our
best estimate of composition and a Bayesian ``standard error'' for
the estimate.

@@

The statistical model treats the observations as having a multinomial
distribution with a (latent) vector of proportions for each grid cell.
\[
y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i}))
\]
where $y_{i}$ is the vector of counts for the $P$ taxa at the $i$th
location, $n_{i}$ is the number of trees counted in the location,
and $\theta(s_{i})$ is the vector of unknown proportions for those
taxa at that location. Note that we use a standard multinomial distribution
without overdispersion as the set of trees in the dataset is roughly
uniformly sampled across the cells or townships \citep{Gori:etal:2015}.

In what follows we describe the basic model for those states for which
we have raw data on the 8 km grid and in Section \ref{sub:Model-for-township}
we describe the extension of the model to accommodate data aggregated
to the township level.

@@

The critical component of the statistical model is the representation
of $\alpha_{p}(s)$ as a spatial process. This process is a prior
structure that serves to smooth across noise in the observations and
allows for interpolation to locations with no data. Apart from the
sum to one constraints, the taxa are considered to be independent
in the prior, as we did not want to impose any structure that ties
the different taxa together, as any correlation will likely vary across
space.

@@

Our first model is a standard conditional autoregressive (CAR) model
\citep{Bane:etal:2003}. We use a standard form of this model, which
treats the four cardinal neighbors of each grid cell as the neighbors
of the grid cell. The corresponding precision matrix has diagonal
elements, $Q_{ii}$, equal to the number of neighbors for the $i$th
area (i.e., four except for cells on the boundary of the domain),
while $Q_{ik}=-1$ (the negative of a weight of one) when areas $i$
and $k$ are neighbors and $Q_{ik}=0$ when they are not. This gives
the following model for the values of $\alpha_{p}(s_{i})$ collected
as a vector across all of the grid cells, $i=1,\ldots,m$: 
\[
\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-})
\]
The use of the generalized inverse notation indicates that $Q$ is
not full-rank, but is of rank $m-1$; this gives an improper prior
on an implicit overall mean for the process values. This specification
is called an \textit{intrinsic conditional autoregression (ICAR)}
and $Q$ can also be expressed as $D-C$ where $C$ is the $m\times m$
adjacency matrix defining the neighborhood relation of the locations;
that is $(C)_{ik}=1$ if locations $i$ and $k$ are neighbors. The
matrix $D$ is an $m\times m$ diagonal matrix containing the row
sums of matrix $C$ as the diagonal entries $(D)_{ii}={\displaystyle \sum_{k=1}^{m}(C)_{ik}}.$

@@

We consider Gaussian processes in the \matern class, using the following
parameterization of the \matern correlation function as 
\begin{equation}
R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
\end{equation}
where $d$ is Euclidean distance, $\rho$ is the spatial range parameter,
and $\mathcal{K}_{\nu}(\cdot)$ is the modified Bessel function of
the second kind, whose order is the smoothness (differentiability)
parameter, $\nu>0$. $\nu=0.5$ gives the exponential covariance.
For any pair of locations, $R(d)$ defines the correlation of the
process, e.g., $\alpha_{p}(s)$, as a function of the distance between
the locations. Considering all pairs of locations, this defines a
correlation matrix for all locations of interest.

The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
to the Matern-based GPs for $\nu=1$ and $\nu=2$. Our second model
is the Lindgren approximation for Matern-based GPs with $\nu=1$.
To implement the Lindgren model, one modifies the $Q$ matrix defined
previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
neighbors are $-2a$. Those for diagonal neighbors are $2$ and those
for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

@@

To ensure that the $\sigma^{2}$ parameter is equivalent between the
two models, we reparameterize, producing our second model:
\[
\alpha_{p}\sim\mbox{N}(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1})
\]

@@

\noindent The ICAR specification contains a set of hyperparameters
$\sigma_{p}^{2}$ for $p$ = $1,...,P$. Following \citep{Gelm:2006}
we use a uniform distribution on each $\sigma_{p}$ parameter, with
upper bound of {[}check max values{]}. For the SPDE model we also
have parameters $\mu_{p}$, which we give flat, non-informative priors
(truncated at $\pm10$), and $\rho_{p}$ which we give uniform priors
on the interval $(0.1,\exp(5))$.

@@

It is well-known that devising an effective MCMC algorithm for models
with latent Gaussian process(es) and a non-Gaussian likelihood is
difficult (cite recent cent/noncent papers). To develop an algorithm,
we make use of a latent variable representation for the multinomial
probit model \citep{McCu:Ross:1994}. The representation introduces
latent variables that allow one to develop a MCMC sampling strategy
that takes advantage of closed form full conditional distributions
(so-called Gibbs sampling steps) for $\alpha_{p}$.

@@

\noindent Consider the following example with two locations that are
neighbors and $P=2$ categories. Each tree $j$ at location $i$ is
associated with two variables $W_{ij1}$ and $W_{ij2}$, governed
by the latent variables $\alpha_{i1}$ and $\alpha_{i2}$, respectively.
Suppose that $\alpha_{i1}>\alpha_{i2}$ for a given location $i$.
Then this model implies that any tree $j$ is more likely to be labeled
1 than 2 at location $i$. The difference between $\alpha_{i1}$ and
$\alpha_{i2}$ explains the \textit{difference} in probability of
\textit{categories} 1 and 2 at location $i$, and the similarity between
$\alpha_{1p}$ and $\alpha_{2p}$ explains the \textit{correlation}
between the probabilities at \textit{locations} 1 and 2 for category
$p$.

@@

We developed an extension of the model described in previous sections
to account for data at a different aggregation than our core 8 km
grid. This extension introduces a new set of latent variables, one
per tree, that indicate the grid cell in which the tree is located.
These latent 'membership' variables, $c_{tj}$, for tree $j$ in township
$t$ can be sampled within the MCMC as additional unknown parameters.
The prior for $c_{tj}$ is a discrete distribution that puts mass
proportional to the overlap between the township in which the tree
is located and the grid cells that intersect the township, $\psi_{1(t)},\ldots,\psi_{T(t)}$,
with the priors across different trees in a township being independent.
\[
c_{tj}\sim\mbox{Multinom}(1,\{\psi_{1(t)},\ldots\psi_{T(t)}\}).
\]

In updating the other parameters in the model, we condition on the
current values, $c_{tj}$, which provides a ``soft'' assignment
of trees to grid cells that respects both the township in which the
tree was sampled and the uncertainty in which grid cell the tree was
sampled.

@@

The \cite{McCu:Ross:1994} representation is convenient for MCMC sampling,
particularly in this high-dimensional spatial context, as it allows
us to draw from the posterior conditional distribution of the $W_{ijp}$
variables (these distributions are truncated normal) in closed form
and to draw the entire vector of latent process values, $\alpha_{p}$,
as a single sample that respects the spatial dependence structure
for each taxon.

>>>>

\title{Statistically-estimated tree composition for the northeastern United
States \\
at the time of European settlement}
\maketitle
\begin{quote}
Christopher J. Paciorek 1{*}, Simon J. Goring 2\&, Andrew Thurman
3\&, Charles V. Cogbill 4, \\
John W. Williams 3,5, David J. Mladenoff 6, Jody A. Peters 7, Jun
Zhu 8, \\
and Jason S. McLachlan 7 \\
{[}author order to be discussed{]}\\
{[}please tell me if you do/don't want your middle initial{]}

{[}let me know if there are any issues with how I've listed your affiliation{]}

1 Department of Statistics, University of California, Berkeley, California,
USA

2 Department of Geography, University of Wisconsin, Madison, Wisconsin,
USA

3 Department of Statistics, University of Iowa, Iowa City, Iowa, USA

4 Harvard Forest, Harvard University, Petersham, Massachusetts, USA

5 Center for Climatic Research, University of Wisconsin, Madison,
Wisconsin, USA

6 Department of Forestry and Wildlife Ecology, University of Wisconsin,
Madison, Wisconsin, USA

7 Department of Biological Sciences, University of Notre Dame, Notre
Dame, Indiana, USA

8 Department of Statistics, University of Wisconsin, Madison, Wisconsin,
USA

{*} Corresponding author

E-mail paciorek@stat.berkeley.edu (CJP)

\& These authors contributed equally to this work. 
\end{quote}
TODO:

@@

\begin{abstract}
We present a data product of the estimated composition of tree taxa
at the time of European settlement of the northeastern United States
and the statistical methodology used to produce the product. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height in 22 taxonomic groupings, generally
at the genus level. The data come from settlement survey records that
provide raw data that are transcribed and then aggregated spatially,
giving count data. The domain is divided into two regions, eastern
(Maine to Ohio) and western (Indian to Minnesota). Public Land Survey
point data in the western region are aggregated to a regular 8 km
grid, while data in the eastern region, from Town Proprietor Surveys,
is aggregated at the township level in irregularly-shaped local administrative
units. The product is based on a Bayesian statistical model fit to
the count data that estimates composition on a regular 8 km grid.
The statistical model allows us to estimate composition at locations
with no data and to smooth over noise caused by limited counts in
locations with data. Critically, it also allows us to quantify uncertainty
in our composition estimates. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale European settlement. In addition to specific
regional questions, the data product can also serve as a baseline
against which to investigate how forests and ecosystems change after
intensive settlement.
\end{abstract}

@@

Historical datasets provide critical context to understand forest
ecology. They allow researchers to define 'baseline' conditions for
conservation management, to understand ecosystem processes at decadal
and centennial scales, and, particularly in regions with widespread
land use change, to understand the extent to which forests after conversion
and regeneration differ from the original forest cover.

Euro-American settlement and subsequent land use change occured in
a time transient fashion across North America, and land surveys that
provide vegetation information followed a similar pattern. Early surveys
(from 1620 until 1825) in the northeastern United States provide areally-aggregated
data at the township level \citep{cogbill2002forests,thompson2013four};
we refer to these as the Town Proprietor Survey (TPS). These surveys
provide composition of major tree species using sometimes inconsistent
common names. Later surveys, after the establishment of the U.S. Public
Land Survey System (PLS) - from 1832 to 1907 - provide point-level
data along a regular grid, with one mile spacing \citet{bourdo1956review,schulte2001original,goring2015composition},
again with sometimes inconsistent common names. Survey instructions
during the PLS varied through time and by point type. This requires
the application of spatially-varying correction factors \citep{cogbill2015corrections,goring2015composition}
to accurately assess stem density, basal area and biomass from the
early settlement records but has limit impact when estimating composition.

Logging, agriculture, and abandonment have left an indelible mark
on forests in the northeastern United States \citet{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
However most studies have assessed these effects in individual states
or smaller domains \citep{rhemtulla2009historical,friedman2005regional}
and with varying spatial resolution, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
provide a dataset for the upper Midwest that is resolved to an 8 x
8 km grid size, providing broad spatial coverage at a spatial scale
that can be compared to the Forest Inventory and Analysis products
\citet{gray2012forest}. Combined with additional, coarsely-sampled
PLS data from Illinois and Indiana and with the TPS data, this gives
us raw data for much of the northeastern United States. However, there
are several limitations of using the raw data that can be alleviated
by the use of a statistical model to develop a statistically-estimated
data product. First, the PLS and TPS data provide only estimates of
within-cell variance that do not account for information from nearby
locations. Second, the available digitized data from Illinois and
Indiana represent a small fraction of those states and missing townships
are common in the TPS data.

Properly assessing uncertainty in ecological data is imperative to
understanding and modelling ecological processes \citep{cressie2009accounting}.
In this way, a model that can account for the spatial structure of
the underlying PLS and TPS data, and provide reliable estimates of
uncertainty across the northeastern United States, provides a valuable
tool for researchers interested in the ecological structure and function
of forests at longer time scales.

@@

The raw data were obtained from survey records collated from across
the northeastern U.S. by a number of researchers. For the states of
Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
subdomain), data are available at PLS survey point locations and have
been aggregated to a regular 8 km grid in the Albers projection. (Note
that for Indiana and Illinois, at the moment trees are associated
with township centroids and then assigned to 8 km grid cells based
on the centroid but in the near future we will have point locations
available for each tree.) For the states of Ohio, Pennsylvania, New
Jersey, New York and the six New England states (the eastern subdomain),
data are aggregated at the township level. There are also data from
a single township in Quebec and a single township in northern Delaware.
Data are essentially complete in Minnesota, Wisconsin and Michigan,
but data in Illinois and Indiana represent a sample of the full set
of grid cells, with survey record transcription ongoing. Data for
the remaining states are available for a subset of the full set of
townships covering the domain. Fig. \ref{fig:domain} {[}Fig. 1 -
Lyx is having a numbering problem{]} shows the domain, indicating
the grid cells and townships with data.

@@

\caption{Spatial domain, with locations with data shown in gray. Locations
are grid cells in western portion and townships in eastern portion.
In addition to locations without data being indicated in white, grid
cells completely covered in water are white (e.g., a few locations
in Minnesota and Wisconsin).}
\end{figure}

Note that surveys occurred over a period of more than 200 years as
European colonists (before U.S. independence) and the United States
settled what is now the northeastern United States. Our estimates
are for the period of settlement represented by the survey data and
therefore are time-transgressive; they do not represent any single
point in time across the domain, but rather the state of the landscape
at the time just before settlement occurred \citep{Whit:1996,Cogb:etal:2002}.

Extensive details on the data are available in \cite{Gori:etal:2015}
and the raw data are available at XXX {[}ask Simon/Charlie about public
access to Paleon composition data{]}. The aggregation into taxonomic
groups is primarily at the genus level, but is at the species level
in some cases of monospecific genera. We model the following 22 taxa
plus an ``other hardwood'' category: Atlantic white cedar (\emph{Chamaecyparis
thyoides}), Ash (\emph{Fraxinus spp.}), Basswood (\emph{Tilia americana}),
Beech (\emph{Fagus grandifolia)}, Birch (\emph{Betula spp.}), Black
gum/sweet gum (\emph{Nyssa sylvatica} and \emph{Liquidambar styraciflua}),
Cedar/juniper (\emph{Juniperus virginiana} and \emph{Thuja occidentalis}),
Cherry (\emph{Prunus spp.}), Chestnut (\emph{Castanea dentata}), Dogwood
(\emph{Cornus spp.}), Elm (\emph{Ulmus spp.}), Fir (\emph{Abies spp.}),
Hemlock (\emph{Tsuga canadensis}), Hickory (\emph{Carya spp.}), Ironwood
(\emph{Carpinus caroliniana} and \emph{Ostrya virginiana}), Maple
(\emph{Acer spp.}), Oak (\emph{Quercus spp.}), Pine (\emph{Pinus spp.}),
Poplar/tulip poplar (\emph{Populus spp.} and \emph{Liriodendron tulipifera}),
Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), Walnut
(\emph{Juglans nigra}). Note that in several cases (e.g., black gum/sweet
gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
in the common tree names used by surveyors, a group represents trees
from different families and even orders. For the western subdomain
we do not fit statistical models for Atlantic white cedar and chestnut
as these have 0 and 6 trees present, respectively. The taxa grouped
into the other hardwood category are those for which fewer than roughly
2000 trees were present in the dataset; however we include Atlantic
white cedar explicitly despite it only having 336 trees in the dataset.
Given the nature of the data, there is not a firm cutoff on the sizes
of trees included, but the data roughly represent trees greater than
8 inches (\textasciitilde{}20 cm) diameter at breast height (dbh),
but with some trees as small as 4 inches. {[}Charlie,Simon, please
check this wording{]} In Indiana and Illinois there are also 402 trees
and 29 trees, respectively, that are between 1 and 4 inches dbh, but
these will be excluded in the next version of the product.

@@

Our domain is a rectangle covering all of the states, in the Albers
projection (NAD 1983 Albers Great Lakes and St. Lawrence), with the
rectangle split into 8 km cells, arranged in a 296 by 180 grid of
cells, with the centroid of the cell in the southeast corner located
at (-71000, 58000). For the modeling of the western subdomain we use
the western-most 146 by 180 grid of cells. For the modeling of the
eastern subdomain we use the eastern-most 180 by 180 grid of cells
and then omit 23 cells in the north and 17 cells in the south outside
the states containing data.

@@

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using full population of all the trees in
an areal region.
\end{enumerate}
The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition in the 23 taxonomic groupings at each of the grid
cells. These samples are the data product (described further in the
Data Product section), and can then be used in subsequent analyses.
The mean and standard deviation of the samples for each cell by taxon
pair represent our best estimate (i.e., prediction) of composition
and a Bayesian ``standard error'' quantifying the uncertainty in
the estimate.

@@

We start by describing the basic model for those states for which
we have raw data on the 8 km grid and in Section \ref{sub:Model-for-township}
we describe the extension of the model to accommodate data aggregated
at the township level.

The statistical model treats the observations as coming from a multinomial
distribution with a (latent) vector of proportions for each grid cell.
\[
y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i}))
\]
where $y_{i}$ is the vector of counts for the $P$ taxa at the $i$th
cell, $n_{i}$ is the number of trees counted in the cell, and $\theta(s_{i})$
is the vector of unknown proportions for those taxa at that cell.
Note that we use a standard multinomial distribution without overdispersion
as the set of trees in the dataset is roughly uniformly sampled across
the cells or townships \citep{Gori:etal:2015}.

@@

The critical component of the statistical model is the representation
of $\alpha_{p}(s)$ as a spatial process. This process is a prior
structure that serves to smooth across noise in the observations and
allows for interpolation to locations with no data. Apart from the
sum to one constraints, the taxa are considered to be independent
in the prior. We did not want to impose any structure that ties the
different taxa together, as any correlation will likely vary across
space.

@@

Our first model is a standard conditional autoregressive (CAR) model
\citep{Bane:etal:2003}. We use a standard form of this model, which
treats the four cardinal neighbors of each grid cell as the neighbors
of the grid cell. The corresponding precision matrix has diagonal
elements, $Q_{ii}$, equal to the number of neighbors for the $i$th
area (i.e., four except for cells on the boundary of the domain),
while $Q_{ik}=-1$ (the negative of a weight of one) when areas $i$
and $k$ are neighbors and $Q_{ik}=0$ when they are not. This gives
the following model for the values of $\alpha_{p}(s_{i})$ collected
as a vector across all of the grid cells, $i=1,\ldots,m$: 
\[
\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-})
\]
The use of the generalized inverse notation indicates that $Q$ is
not full-rank, but is of rank $m-1$; this gives an improper prior
on an implicit overall mean for the process values. This specification
is called an \textit{intrinsic conditional autoregression (ICAR)}
and we can write $Q=D-C$ where $C$ is the $m\times m$ adjacency
matrix defining the neighborhood relation of the locations; that is
$(C)_{ik}=1$ if locations $i$ and $k$ are neighbors. The matrix
$D$ is an $m\times m$ diagonal matrix containing the row sums of
matrix $C$ as the diagonal entries $(D)_{ii}={\displaystyle \sum_{k=1}^{m}(C)_{ik}}.$

@@

We consider Gaussian processes in the \matern class, using the following
parameterization of the \matern correlation function as 
\begin{equation}
R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
\end{equation}
where $d$ is Euclidean distance, $\rho$ is the spatial range parameter,
and $\mathcal{K}_{\nu}(\cdot)$ is the modified Bessel function of
the second kind, whose order is the smoothness (differentiability)
parameter, $\nu>0$. $\nu=0.5$ gives the exponential covariance.
For any pair of locations, $R(d)$ defines the correlation of the
process, (i.e., $\alpha_{p}(s)$ in our context), as a function of
the distance between the locations. Considering all pairs of locations,
this defines a correlation matrix for all locations of interest.

The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
to the Matern-based GP for $\nu=1$ and $\nu=2$. Our second model
is the Lindgren approximation for Matern-based GPs with $\nu=1$.
To implement the Lindgren model, one modifies the $Q$ matrix defined
previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

@@

To ensure that the $\sigma^{2}$ parameter is mathematically equivalent
between the two models, we reparameterize, producing our second model:
\[
\alpha_{p}\sim\mbox{N}\left(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1}\right)
\]

@@

\noindent The ICAR specification contains a set of hyperparameters
$\sigma_{p}^{2}$ for $p$ = $1,...,P$. Following \cite{Gelm:2006}
we use a uniform distribution on each $\sigma_{p}$ parameter, with
upper bound of {[}check max values{]}. For the SPDE model we also
have parameters $\mu_{p}$, which we give flat, non-informative priors
(truncated at $\pm10$), and $\rho_{p}$ which we give uniform priors
on the interval $(0.1,\exp(5))$.

@@

It is well-known that devising an effective MCMC algorithm for models
with latent Gaussian process(es) and a non-Gaussian likelihood is
difficult (cite recent cent/noncent papers \citep{Rue:Held:2005,Chri:etal:2006,Tan:Nott:2013}.
To develop an algorithm, we make use of a latent variable representation
for the multinomial probit model \citep{McCu:Ross:1994}. The representation
introduces latent variables that allow one to develop a MCMC sampling
strategy that takes advantage of closed form full conditional distributions
(so-called Gibbs sampling steps) for $\alpha_{p}$.

@@

\noindent Consider the following example with two locations that are
neighbors and $P=2$ categories. Each tree $j$ at location $i$ is
associated with two variables $W_{ij1}$ and $W_{ij2}$, governed
by the latent variables $\alpha_{1i}$ and $\alpha_{2i}$, respectively.
Suppose that $\alpha_{1i}>\alpha_{2i}$ for a given location $i$.
Then this model implies that any tree $j$ is more likely to be labeled
1 than 2 at location $i$. The difference between $\alpha_{1i}$ and
$\alpha_{2i}$ explains the \textit{difference} in probability of
\textit{categories} 1 and 2 at location $i$, and the similarity between
$\alpha_{p1}$ and $\alpha_{p2}$ explains the \textit{correlation}
between the probabilities at \textit{locations} 1 and 2 for category
$p$.

@@

We developed an extension of the model described in previous sections
to account for data at a different aggregation than our core 8 km
grid. This extension introduces a new set of latent variables, one
per tree, that indicate the grid cell in which the tree is located.
These latent 'membership' variables, $c_{tj}$, for tree $j$ in township
$t$, $t=1,\ldots,T$ can be sampled within the MCMC as additional
unknown parameters. The prior for $c_{tj}$ is a discrete distribution
that puts mass proportional to the areal overlap between the township
in which the tree is located and the $m$ grid cells, with the priors
across different trees in a township being independent and with most
of the . 
\[
c_{tj}\sim\mbox{Multinom}(1,\{\psi_{1},\ldots,\psi_{m}\}).
\]
Because the townships overlap a limited number of grid cells, most
of the $\psi_{1},\ldots,\psi_{m}$ values are zero.

Using the latent variable representation, we have that $W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{c_{tj}},1)$
for tree $j$ in township $t$. In updating the other parameters in
the model during the MCMC (specifically the $\alpha$ values), we
condition on the current values, $c_{tj}$, which provides a ``soft''
assignment of trees to grid cells that respects both the known township
in which the tree occurred and the uncertainty in which grid cell
the tree occurred.

@@

The \cite{McCu:Ross:1994} representation is convenient for MCMC sampling,
particularly in this high-dimensional spatial context, as it allows
us to draw from the posterior conditional distributions of the $W_{ijp}$
variables (these distributions are truncated normal) in closed form
and to draw the entire vector of latent process values for each taxon,
$\alpha_{p}$, as a single sample that respects the spatial dependence
structure for each taxon.

While the latent variable representation provides great advantages
in the MCMC sampling for each $\alpha_{p}$ compared to joint Metropolis
updates or updating each location individually, there is still strong
dependence between the hyperparameters, $\{\sigma_{p}^{2},\mu_{p},\rho_{p}\}$
and the latent process values (as well as between the latent process
values and the latent $W_{ijp}$ variables). To address the first,
we developed a 'cross-level' joint updating strategy in which we propose
$\phi\in\{\sigma_{p}\},p=1,\ldots,P$ (and for the SPDE model, $\phi\in\{\{\sigma_{p}\},\{\mu_{p}\},\{\rho_{p}\}\}),$
via a Metropolis-style random walk and then conditional on the proposed
value of $\phi^{*}$ propose $\alpha_{p}$ from its full conditional
distribution given $\phi^{*}$and the latent $W$ variables. This
is equivalent to sampling from the marginalized (with respect to $\alpha_{p}$)
distribution of $\phi$ conditional on $W_{p}$. Note that in sampling
$\sigma_{p}$and $\rho_{p}$ in the SPDE model, for each $p$, we
jointly propose $\{\sigma_{p},\rho_{p}\}$ using an adaptive Metropolis
proposal and then use the cross-level strategy to also propose $\alpha_{p}$
as part of a single accept-reject step. For these various joint samples
of hyperparameters and $\alpha_{p}$, we use adaptive Metropolis sampling
\citep{Shab:Well:2011}.
